Highly Available Transactions: Virtues and Limitations\iftechreport\\(Extended Version)\fi

### 2013-10-05 writeup/evaluation.tex
Grammar

\vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C
transactions are well-served by HATs. The two problematic
transactions, New-Order and Payment, rely on non-monotonic state
update. The former can be modified to ensure ID uniqueness but not
sequential ID ordering, while the latter is inherently non-monotonic,
requiring external compensation or stronger consistency
protocols. Based on these experiences and discussions with
practitioners, we believe that HAT guarantees can provide useful
semantics for a large class of application functionality, while a
(possibly small) subset of operations will require stronger,
unavailable properties.

@@

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters and stick all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we deploy 5 Amazon EC2
\texttt{m1.xlarge} instances as servers in each cluster. For our
workload, we link our client library to the YCSB
benchmark~\cite{ycsb}, which is well-suited to LevelDB's key-value
schema, grouping every eight YCSB operations from the default workload
(50\% reads, 50\% writes) to form a transaction. We increase the
number of keys in the workload from the default 1,000 to 100,000 with
uniform random key access, keeping the default value size of $1KB$,
and running YCSB for 180 seconds per configuration.

@@

In contrast, when the two clusters are deployed across the continental
United States (Figure~\ref{fig:wan-exp}B), the average latency of
\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
increase; average $37$ms latency per operation). For the same number
of YCSB client threads, \texttt{master} has substantially lower
throughput than the HAT configurations. Increasing the number of YCSB
clients \textit{does} increase the throughput of \texttt{master}, but
our Thrift-based server-side connection processing did not gracefully
handle more than several thousand concurrent connections. In contrast,
across two datacenters, the performance of eventual, RC, and MAV are
near-identical to a single-datacenter deployment.

@@

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for \texttt{eventual}, \texttt{RC}, and \texttt{MAV}. \texttt{RC} and
\texttt{eventual} scale perfectly with an approximately $5$x
throughput increase moving from $5$ to $25$ servers per cluster. For
the same configuration, \texttt{MAV} scales by $3.8$x, achieving over
$260,000$ operations per second. \texttt{MAV} performance is due to
resource contention---with a memory-backed database (instead of
LevelDB), \texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
performance heterogeneity across servers (Calvin's authors report
similar heterogeneity on EC2~\cite{calvin}). Our early experiments
with a newer database prototype with more efficient (non-Thrift) RPC
and connection pooling suggest that this scalability can be
substantially reduced.

>>>>

\vspace{.5em}\noindent\textit{Summary.} Many---but not all---TPC-C
transactions are well served by HATs. The two problematic
transactions, New-Order and Payment, rely on non-monotonic state
update. The former can be modified to ensure ID uniqueness but not
sequential ID ordering, while the latter is inherently non-monotonic,
requiring external compensation or stronger consistency
protocols. Based on these experiences and discussions with
practitioners, we believe that HAT guarantees can provide useful
semantics for a large class of application functionality, while a
(possibly small) subset of operations will require stronger,
unavailable properties.

@@

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters and stick all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we deploy 5 Amazon EC2
\texttt{m1.xlarge} instances as servers in each cluster. For our
workload, we link our client library to the YCSB
benchmark~\cite{ycsb}, which is well suited to LevelDB's key-value
schema, grouping every eight YCSB operations from the default workload
(50\% reads, 50\% writes) to form a transaction. We increase the
number of keys in the workload from the default 1,000 to 100,000 with
uniform random key access, keeping the default value size of $1KB$,
and running YCSB for 180 seconds per configuration.

@@

In contrast, when the two clusters are deployed across the continental
United States (Figure~\ref{fig:wan-exp}B), the average latency of
\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
increase; average $37$ms latency per operation). For the same number
of YCSB client threads, \texttt{master} has substantially lower
throughput than the HAT configurations. Increasing the number of YCSB
clients \textit{does} increase the throughput of \texttt{master}, but
our Thrift-based server-side connection processing did not gracefully
handle more than several thousand concurrent connections. In contrast,
across two datacenters, the performance of eventual, RC, and MAV are
near identical to a single-datacenter deployment.

@@

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for \texttt{eventual}, \texttt{RC}, and \texttt{MAV}. \texttt{RC} and
\texttt{eventual} scale linearly: increasing the number of servers per
cluster from $5$ to $25$ yields an approximately $5$x throughput
increase. For the same configurations, \texttt{MAV} scales by $3.8$x,
achieving over $260,000$ operations per second. \texttt{MAV} suffers
from contention in LevelDB---with a memory-backed database,
\texttt{MAV} scales by $4.25$x (not shown)---and MAV-related
performance heterogeneity across servers (Calvin's authors report
similar heterogeneity on EC2~\cite{calvin}). Initial experiments with
a newer prototype including more efficient (non-Thrift) RPC and
connection pooling suggest that this scalability can be substantially
improved.

### 2013-10-05 writeup/hat-analysis.tex
Grammar

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}. In this model, writes to each object are totally
ordered, corresponding to the order in which they are installed in the
database. In a distributed database, different replicas may receive
writes to their local copies of data at different times but should
handle concurrent updates (i.e., overwrites) in accordance with the
total order for each item. \textit{PL-1} requires that writes to
different objects be ordered consistently across transactions,
prohibiting Adya's phenomenon $G0$ (also called ``Dirty
Writes''~\cite{ansicritique}). If we build a graph of transactions
with edges from one transaction to another and, when the former
overwrites the latter's write to the same object, then, under Read
Uncommitted, the graph should not contain cycles~\cite{adya}. Consider
the following example:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\end{align*}
In this example, under Read Uncommitted, it is impossible for the
database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
achieved by marking each of a transactions' writes with the same
timestamp (unique across transactions; e.g., combining a client's ID
with a sequence number) and applying a ``last writer wins'' conflict
reconciliation policy at each replica. Later properties will
strengthen Read Uncommitted.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default of many DBMSs (as in
Section~\ref{sec:modernacid}). Centralized implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. These implementations often provide recency and monotonicity
properties beyond what is implied by the name ``Read Committed'' and
what is captured by the implementation-agnostic definition: under
Read Committed, transactions should not access uncommitted or
intermediate versions of data items. This prohibits both ``Dirty
Writes'', as above, and also ``Dirty Reads'' phenomena.  This
isolation is Adya's \textit{PL-2} and is formalized by prohibiting
Adya's \textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2]
from Berenson et al.). For instance, in the example below, $T_3$
should never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read
$a=3$:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-1em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of data, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. Unlike a lock-based implementation, this implementation
does not provide recency or monotonicity guarantees but it satisfies
the implementation-agnostic definition.

Several different properties have been labeled \textbf{Repeatable
  Read} isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures the spirit of the term: if a
transaction reads the same data more than once, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation}, and, if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
the stronger property of \textbf{Predicate Cut-Isolation}. In the
example below, under both levels of cut isolation, $T_3$ must read
$a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having transactions store a copy of any read data at the client such
that the values that they read for each item never changes unless they
overwrite it themselves. These stored values can be discarded at the
end of each transaction and can alternatively be accomplished on
(sticky) replicas via multi-versioning. Predicate Cut Isolation is
also achievable in HAT systems via similar caching middleware or
multi-versioning that track entire logical ranges of predicates in
addition to item-based reads.

@@

Perplexingly, discussions of MAV are absent from existing treatments of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), MAV is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore,
  kemme-thesis}. In contrast, in a distributed environment, MAV over
arbitrary groups of non-colocated items is considerably more difficult
to achieve with high availability.

@@

As a straw-man, replicas can store all versions ever written to each
data item. Replicas can gossip information about versions they have
observed and construct a lower bound on the versions that can be found
on every replica (which can be represented by either a list of
versions, or, more realistically, a vector clock). At the start of
each transaction, clients can choose a \textit{read timestamp} that is
lower than or equal to the this global lower bound, and, during
transaction execution, replicas return the latest version of each item
that is not greater than the client's chosen timestamp. If this lower
bound is advanced along transactional boundaries, clients will observe
MAV. This algorithm has several variants in the
literature~\cite{readonly, swift}: bounds can be represented by vector
clocks, and older versions can be asynchronously garbage collected.

@@

\vspace{.5em}\noindent\textbf{{Monotonic writes}} requires that each
session's writes become visible in the order they were submitted by
the client. Any order on transactions should also be consistent with
any precedence that a global observer would see.

@@

The above guarantees can be achieved by forcing servers to wait to
reveal new writes (say, by buffering them in separate local storage)
until each write's respective dependencies are visible on all
replicas. This mechanism effectively ensures that all clients read
from a globally agreed upon lower bound on the versions written. This
is highly available as a client will never block due to inability to
find a server with a sufficiently up-to-date version of a data
item. However, it does not imply that transactions will read their own
writes or, in the presence of partitions, make forward progress
through the version history. The problem is that under non-sticky
availability, a system must handle the possibility that, under a
partition, an unfortunate client will be forced to issue its next
requests against a partitioned, out-of-date server.

>>>>

To begin, Adya captures \textbf{Read Uncommitted} isolation as
\textit{PL-1}. In this model, writes to each object are totally
ordered, corresponding to the order in which they are installed in the
database. In a distributed database, different replicas may receive
writes to their local copies of data at different times but should
handle concurrent updates (i.e., overwrites) in accordance with the
total order for each item. \textit{PL-1} requires that writes to
different objects be ordered consistently across transactions,
prohibiting Adya's phenomenon $G0$ (also called ``Dirty
Writes''~\cite{ansicritique}). If we build a graph of transactions
with edges from one transaction to another and, when the former
overwrites the latter's write to the same object, then, under Read
Uncommitted, the graph should not contain cycles~\cite{adya}. Consider
the following example:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\end{align*}
In this example, under Read Uncommitted, it is impossible for the
database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$. Read Uncommitted is easily
achieved by marking each of a transaction's writes with the same
timestamp (unique across transactions; e.g., combining a client's ID
with a sequence number) and applying a ``last writer wins'' conflict
reconciliation policy at each replica. Later properties will
strengthen Read Uncommitted.

\textbf{Read Committed} isolation is particularly important in
practice as it is the default isolation level of many DBMSs
(Section~\ref{sec:modernacid}). Centralized implementations differ,
with some based on long-duration exclusive locks and short-duration
read locks~\cite{gray-isolation} and others based on multiple
versions. These implementations often provide recency and monotonicity
properties beyond what is implied by the name ``Read Committed'' and
what is captured by the implementation-agnostic definition: under Read
Committed, transactions should not access uncommitted or intermediate
versions of data items. This prohibits both ``Dirty Writes'', as
above, and also ``Dirty Reads'' phenomena.  This isolation is Adya's
\textit{PL-2} and is formalized by prohibiting Adya's
\textit{G1\{a-c\}} (or ANSI's $P1$, or ``broad'' $P1$ [2.2] from
Berenson et al.). For instance, in the example below, $T_3$ should
never see $a=1$, and, if $T_2$ aborts, $T_3$ should not read $a=3$:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\vspace{-1em}
\end{align*}
It is fairly easy for a HAT system to prevent ``Dirty Reads'': if each
client never writes uncommitted data to shared copies of data, then
transactions will never read each others' dirty data. As a simple
solution, clients can buffer their writes until they commit, or,
alternatively, can send them to servers, who will not deliver their
value to other readers until notified that the writes have been
committed. Unlike a lock-based implementation, this implementation
does not provide recency or monotonicity guarantees but it satisfies
the implementation-agnostic definition.

Several different properties have been labeled \textbf{Repeatable
  Read} isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures the spirit of the term: if a
transaction reads the same data more than once, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation}, and, if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
the stronger property of \textbf{Predicate Cut-Isolation}. In the
example below, under both levels of cut isolation, $T_3$ must read
$a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having transactions store a copy of any read data at the client such
that the values that they read for each item never changes unless they
overwrite it themselves. These stored values can be discarded at the
end of each transaction and can alternatively be accomplished on
(sticky) replicas via multi-versioning. Predicate Cut Isolation is
also achievable in HAT systems via similar caching middleware or
multi-versioning that track entire logical ranges of predicates in
addition to item based reads.

@@

Perplexingly, discussions of MAV are absent from existing treatments of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), MAV is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore,
  kemme-thesis}. In contrast, in a distributed environment, MAV over
arbitrary groups of non-co-located items is considerably more difficult
to achieve with high availability.

@@

As a straw man, replicas can store all versions ever written to each
data item. Replicas can gossip information about versions they have
observed and construct a lower bound on the versions that can be found
on every replica (which can be represented by either a list of
versions, or, more realistically, a vector clock). At the start of
each transaction, clients can choose a \textit{read timestamp} that is
lower than or equal to the this global lower bound, and, during
transaction execution, replicas return the latest version of each item
that is not greater than the client's chosen timestamp. If this lower
bound is advanced along transactional boundaries, clients will observe
MAV. This algorithm has several variants in the
literature~\cite{readonly, swift}, and older versions can be
asynchronously garbage collected.

@@

\vspace{.5em}\noindent\textbf{{Monotonic writes}} requires that each
session's writes become visible in the order they were submitted. Any
order on transactions should also be consistent with any precedence
that a global observer would see.

@@

The above guarantees can be achieved by forcing servers to wait to
reveal new writes (say, by buffering them in separate local storage)
until each write's respective dependencies are visible on all
replicas. This mechanism effectively ensures that all clients read
from a globally agreed upon lower bound on the versions written. This
is highly available because a client will never block due to inability
to find a server with a sufficiently up-to-date version of a data
item. However, it does not imply that transactions will read their own
writes or, in the presence of partitions, make forward progress
through the version history. The problem is that under non-sticky
availability, a system must handle the possibility that, under a
partition, an unfortunate client will be forced to issue its next
requests against a partitioned, out-of-date server.

### 2013-10-05 writeup/intro.tex
Grammar

At a high level, the virtues of HATs are guaranteed responses from any
replica, low latency, and a range of semantic guarantees including
several whose usefulness is widely acepted such as Read
Committed. However, highly available systems are fundamentally unable
to prevent concurrent updates to shared data items and cannot provide
recency guarantees for reads. To understand when these virtues and
limitations are relevant in practice, we survey both practitioner
accounts and academic literature, perform experimental analysis on
modern cloud infrastructure, and analyze representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints for multi-item updates and, in some cases, provide limited
uniqueness guarantees. However, HATs can fall short for applications
with concurrency-sensitive operations, requiring non-HA coordination.

Finally, we recognize that the large variety of ACID isolation levels
and distributed consistency models (and therefore those in our
taxonomization) can be confusing; the subtle distinctions between
models may appear to be of academic concern. Accordingly, we offer the
following pragmatic takeaways:
\begin{introenumerate}
\item The default (and sometimes strongest) configurations of most
  widely-deployed database systems expose a range of anomalies that
  can compromise application-level consistency.

\item Many of these ``weak isolation'' models are achievable without
  sacrificing high availability if implemented correctly. However,
  none of these models prevents concurrent modifications.

>>>>

At a high level, the virtues of HATs are guaranteed responses from any
replica, low latency, and a range of semantic guarantees including
several whose usefulness is widely accepted such as Read
Committed. However, highly available systems are fundamentally unable
to prevent concurrent updates to shared data items and cannot provide
recency guarantees for reads. To understand when these virtues and
limitations are relevant in practice, we survey both practitioner
accounts and academic literature, perform experimental analysis on
modern cloud infrastructure, and analyze representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints for multi-item updates and, in some cases, provide limited
uniqueness guarantees. However, HATs can fall short for applications
with concurrency-sensitive operations, requiring non-HA coordination.

Finally, we recognize that the large variety of ACID isolation levels
and distributed consistency models (and therefore those in our
taxonomy) can be confusing; the subtle distinctions between models may
appear to be of academic concern. Accordingly, we offer the following
pragmatic takeaways:
\begin{introenumerate}
\item The default (and sometimes strongest) configurations of most
  widely deployed database systems expose a range of anomalies that
  can compromise application-level consistency.

\item Many of these ``weak isolation'' models are achievable without
  sacrificing high availability if implemented correctly. However,
  none of the achievable models prevents concurrent modifications.

### 2013-10-05 writeup/model.tex
Grammar

Traditionally, a system provides {\textbf{high availability}} if every
user that can contact a correct (non-failing) server eventually
receives a response from that server, even in the presence of
arbitrary, indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition from the
  distributed systems literature, systems that require a majority of
  servers to be online are not available. Similarly, a system which
  guarantees that servers provide a response with high probability is
  not available. This admittedly stringent requirement matches the
  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
distributed database, designated servers might perform operations for
different data items; a server that can handle an operation for a
given data item is called a \textit{replica} for that
item.\footnote{There is a further distinction between a \textit{fully
    replicated} system, in which all servers are replicas for all data
  items and a \textit{partially replicated} system, in which at least
  one server acts as a replica for a proper subset of all data
  items. For generality, and, given the prevalence of these
  ``sharded'' or ``partitioned'' systems~\cite{ bigtable, pnuts,
    spanner, dynamo, hstore}, we consider partial replication here.}

@@

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical replica(s) across subsequent
operations, whereby each of the client's prior operations (but not
necessarily other clients' operations) are reflected in the database
state that they observe. As we will discuss in Section~\ref{sec:hats},
clients can ensure continuity between operations (e.g., reading their
prior updates to a data item) by maintaining affinity or
``stickiness'' with a server or set of servers~\cite{vogels-defs}. In
a fully-replicated system, where all servers are replicas for all data
items, stickiness is simple: a client can maintain stickiness by
contacting the same server for each of its requests. However, to stay
``sticky'' in a partially-replicated system, where servers are
replicas for subsets of the set of data items (which we consider in
this paper), a client must maintain stickiness with a single
\textit{logical} copy of the database, which may consist of multiple
physical servers. We say that a system provides \textbf{sticky
  availability} if, whenever a client's transactions is executed
against a copy of database state that reflects all of the client's prior
transactions, it eventually receives a response, even in the presence
of indefinitely long partitions. A client may choose to become sticky
available by acting as a server itself; for example, a client might
cache its reads and writes~\cite{bolton, sessionguarantees,
  swift}. Any guarantee achievable in a highly available system is
achievable in a sticky high availability system but not vice-versa.

>>>>

Traditionally, a system provides {\textbf{high availability}} if every
user that can contact a correct (non-failing) server eventually
receives a response from that server, even in the presence of
arbitrary, indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition from the
  distributed systems literature, systems that require a majority of
  servers to be online are not available. Similarly, a system which
  guarantees that servers provide a response with high probability is
  not available. This admittedly stringent requirement matches the
  assumptions made in the CAP Theorem~\cite{gilbert-cap} and
  guarantees low latency~\cite{abadi-pacelc}.} As in a standard
distributed database, designated servers might perform operations for
different data items. A server that can handle an operation for a
given data item is called a \textit{replica} for that
item.\footnote{There is a further distinction between a \textit{fully
    replicated} system, in which all servers are replicas for all data
  items and a \textit{partially replicated} system, in which at least
  one server acts as a replica for a proper subset of all data
  items. For generality, and, given the prevalence of these
  ``sharded'' or ``partitioned'' systems~\cite{ bigtable, pnuts,
    spanner, dynamo, hstore}, we consider partial replication here.}

@@

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical replica(s) across subsequent
operations, whereby each of the client's prior operations (but not
necessarily other clients' operations) are reflected in the database
state that they observe. As we will discuss in Section~\ref{sec:hats},
clients can ensure continuity between operations (e.g., reading their
prior updates to a data item) by maintaining affinity or
``stickiness'' with a server or set of servers~\cite{vogels-defs}. In
a fully replicated system, where all servers are replicas for all data
items, stickiness is simple: a client can maintain stickiness by
contacting the same server for each of its requests. However, to stay
``sticky'' in a partially-replicated system, where servers are
replicas for subsets of the set of data items (which we consider in
this paper), a client must maintain stickiness with a single
\textit{logical} copy of the database, which may consist of multiple
physical servers. We say that a system provides \textbf{sticky
  availability} if, whenever a client's transactions is executed
against a copy of database state that reflects all of the client's prior
transactions, it eventually receives a response, even in the presence
of indefinitely long partitions. A client may choose to become sticky
available by acting as a server itself; for example, a client might
cache its reads and writes~\cite{bolton, sessionguarantees,
  swift}. Any guarantee achievable in a highly available system is
achievable in a sticky high availability system but not vice-versa.

### 2013-10-05 writeup/relatedwork.tex
Grammar

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. Recent work on
PACELC expands CAP by considering connections between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. There are a wide range of coordination-avoiding
``optimistic replication'' strategies~\cite{optimistic} and several
recent efforts at further taxonomizing these strategies in light of
current practice and the proliferation of ``eventually consistent''
stores~\cite{bailis-ec, bernstein-survey} (notably, Bernstein and
Das~\cite{bernstein-survey} specifically highlight the importance of
stickiness~\cite{sessionguarantees, vogels-defs}---which we formalize
in Section~\ref{sec:sticky}).  Aside from our earlier workshop paper
discussing transactional availability, real-world ACID, and HAT RC and
I-CI~\cite{hat-hotos}---which this work expands with additional
semantics, algorithms, and analysis---we believe this paper is the
first to explore the connections between transactional semantics, data
consistency, and (Gilbert and Lynch~\cite{gilbert-cap}) availability.

>>>>

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. Recent work on
PACELC expands CAP by considering connections between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. There are a wide range of coordination-avoiding
``optimistic replication'' strategies~\cite{optimistic} and several
recent efforts at further understanding these strategies in light of
current practice and the proliferation of ``eventually consistent''
stores~\cite{bailis-ec, bernstein-survey}. Notably, Bernstein and
Das~\cite{bernstein-survey} specifically highlight the importance of
stickiness~\cite{sessionguarantees, vogels-defs}---which we formalize
in Section~\ref{sec:sticky}.  Aside from our earlier workshop paper
discussing transactional availability, real-world ACID, and HAT RC and
I-CI~\cite{hat-hotos}---which this work expands with additional
semantics, algorithms, and analysis---we believe this paper is the
first to explore the connections between transactional semantics, data
consistency, and (Gilbert and Lynch~\cite{gilbert-cap}) availability.
